MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-08-01 MAESTROeX_I/algorithm.tex
in algorithm description, replace "edge" with "face" where appropriate

There are a few key numerical modules we use in each time step.
\begin{itemize}
\item {\bf Average}$[\phi]\rightarrow[\overline\phi]$ computes the lateral average of a quantity over a layer at constant radius $r$, as described above in Section \ref{Sec:Spatial}.
\item {\bf Enforce HSE}$[\rho_0]\rightarrow[p_0]$ computes the base state pressure, $p_0$, from a base state density, $\rho_0$ by integrating the hydrostatic equilibrium condition in one dimension. 
This follows equation (A10) in Paper V, noting that for the irregularly spaced base state case, $\Delta r$ is not constant, where $\Delta r_{j+1/2} = r_{j+1} - r_j$ for cell-edge $j+1/2$.  
The base state pressure remains equal to a constant value at the location of a prescribed cutoff density outward for the entire simulation.
\item {\bf React State}$[(\rho X_k)^{\rm in},(\rho h)^{\rm in},p_0]\rightarrow[(\rho X_k)^{\rm out},(\rho h)^{\rm out},(\rho\dot\omega),(\rho\Hnuc)]$ 
uses VODE \citep{vode} to integrate the species and enthalpy due to reactions over $\Delta t/2$ by solving
\begin{equation}
\frac{dX_k}{dt} = \dot\omega_k(\rho,X_k,T); \qquad
\frac{dT}{dt} = \frac{1}{c_p}\left(-\sum_k\xi_k\dot\omega_k + H_{\rm nuc})\right)
\end{equation}
The inputs are the species, enthalpy, and base state pressure, and the outputs are the species, enthalpy, reaction rates, and nuclear energy generation rate.
See Paper III for details.
\end{itemize}
Each time step is constrained by the standard advective CFL condition,
\begin{equation}
\Delta t = \sigma^{\rm CFL} \min_i(\Delta x / U_i),
\end{equation}
where for our simulations we typically use $\sigma^{\rm CFL}\sim 0.7$ and the minimum is taken over all spatial directions over all cells.
There are additional constraints on the time step that are typically much less restrictive than the advective CFL including
the acceleration due to the buoyancy force (sometimes in effect when the velocity is approximately zero at the start of some simulations) 
and the local magnitude of the divergence constraint (to prevent too much mass evacuation from a cell in a time step); see Section 3.4 in Paper III for details.

@@

%--------------------------------------------------------------------------
% STEP 3
%--------------------------------------------------------------------------
\item[Step 3] {\em Construct a time-centered advective velocity on edges, $\uadvone$.}

The construction of face-centered time-centered states used to discretize the
advection terms for velocity, species, and enthalpy, are performed using
a standard multidimensional corner transport upwind approach
\citep{colella1990multidimensional,saltzman1994unsplit} with the piecewise-parabolic method (PPM)
one-dimensional tracing \citep{colella1984piecewise}.  The full details of this
Godunov advection approach for all steps in this algorithm are described 
in Appendix A of \cite{XRB_III}.
Here we use equation (\ref{eq:momentum}) to compute time-centered edge velocities, $\uadvonedag$.
The $\dagger$ superscript refers to the fact that the predicted velocity field does not satisfy the divergence constraint,
\begin{equation}
\nabla \cdot \left(\beta_0^n \uadvone\right) = \beta_0^n \left[S^{\nph,\star} - \frac{1}{\gammaonebar^np_0^n}\left(\frac{\partial p_0}{\partial t}\right)^{n-\myhalf} \right].\label{eq:div1}
\end{equation}
 We project $\uadvonedag$ onto the space of velocities that satisfy the constraint to obtain $\uadvone$.
Each projection step in the algorithm involves the solution of a variable-coefficient Poisson solve using multigrid.
Note that we still employ velocity-splitting as described by equation (\ref{eq:velsplit}) for this step 
in order to enforce the appropriate behavior of the system near the edge of the star as determined by the cutoff density.
The details of this ``MAC'' projection are provided in Appendix \ref{Sec:Projection}.

@@

\item Compute the time-centered species edge states, $(\rho X_k)^{\nph,\pred}$,
  for the conservative update of $(\rho X_k)^{(1)}$ using a Godunov approach \citep{XRB_III}.
  As described in Paper V, for robust numerical slope limiting we predict 
  $\rho'^n=\rho^n-\rho_0^n$ and $X_k^n$ to edges
  and here we spatially interpolate $\rho_0^n$ to edges to assemble the fluxes.

@@

\item Compute the time-centered enthalpy edge state, $(\rho h)^{\nph,\pred},$
  for the conservative update of $(\rho h)^{(1)}$ using using a Godunov approach \citep{XRB_III}.
  As described in Paper V, for robust numerical slope limiting 
  we predict $(\rho h)'^n=(\rho h)^n-(\rho h)_0^n$ to edges,
  where $(\rho h)_0^n$ is obtained by calling {\bf Average}$[(\rho h)^n]\rightarrow[(\rho h)_0]$,
  and here we spatially interpolate $(\rho h)_0^n$ to edges to assemble the fluxes.

@@

%--------------------------------------------------------------------------
% STEP 7
%--------------------------------------------------------------------------
\item[Step 7] {\em Construct a time-centered advective velocity on edges, $\uadvtwo$.}

@@

\item Compute the time-centered species edge states, $(\rho X_k)^{\nph,\pred}$,
  for the conservative update of $(\rho X_k)^{(1)}$ using a Godunov approach \citep{XRB_III}.
  Again, we predict $\rho'^n=\rho^n-\rho_0^n$ and $X_k^n$ to edges
  and here we spatially interpolate $(\rho_0^n+\rho_0^{n+1,*})/2$ to edges to assemble the fluxes.

@@

\item Compute the time-centered enthalpy edge state, $(\rho h)^{\nph,\pred},$
  for the conservative update of $(\rho h)^{(1)}$ using a Godunov approach \citep{XRB_III}.
  Again, we predict $(\rho h)'^n=(\rho h)^n-(\rho h)_0^n$ to edges
  and here we spatially interpolate $[(\rho h)_0^n)+(\rho h)_0^{n+1,*}]/2$ to edges to assemble the fluxes.

@@

First, we compute the time-centered edge velocities, $\Ub^{\nph,\pred}$
using a Godunov approach \citep{XRB_III}. Then, we define
\begin{equation}
\rho^\nph = \frac{\rho^n + \rho^{n+1}}{2}, \qquad \rho_0^\nph = \frac{\rho_0^n + \rho_0^{n+1}}{2}.
\end{equation}
We update the velocity field $\Ub^n$ to $\Ub^{n+1,\dagger}$ by discretizing
equation (\ref{eq:momentum}) as
\begin{equation}
\Ub^{n+1,\dagger}
= \Ub^n - \dt \left[\uadvtwo \cdot \nabla \Ub^{\nph,\pred} \right]
 - \dt \left[ \frac{\beta_0^\nph}{\rho^\nph} \nabla \left( \frac{\pi^\nmh}{\beta_0^\nmh} \right) + \frac{\left(\rho^\nph-\rho_0^\nph\right)}{\rho^\nph} g^{\nph} \eb_r \right],
\end{equation}
Again, the $\dagger$ superscript refers
to the fact that the updated velocity does not satisfy the divergence constraint,
\begin{equation}
\nabla \cdot \left(\beta_0^{n+1} \Ub^{n+1} \right) = \beta_0^{n+1} \left[ S^{n+1} - \frac{1}{\gammaonebar^{n+1}p_0^{n+1}}\left(\frac{\partial p_0}{\partial t}\right)^{\nph}\right].\label{eq:div3}
\end{equation}
We use an approximate projection to project $\Ub^{n+1,\dagger}$ onto the space of velocities that satisfy the constraint to obtain $\Ub^{n+1}$ using a ``nodal'' projection.
This projection necessarily differs from the MAC projection used in
{\bf Step 3} and {\bf Step 7} because the velocities in those steps are defined
on edges and $\Ub^{n+1}$ is defined at cell centers, requiring different divergence
and gradient operators.
Furthermore, as part of the nodal projection, we also define a nodal new-time perturbational pressure, $\pi^\nph$.
Refer to Appendix \ref{Sec:Projection} for more details.

>>>>

There are a few key numerical modules we use in each time step.
\begin{itemize}
\item {\bf Average}$[\phi]\rightarrow[\overline\phi]$ computes the lateral average of a quantity over a layer at constant radius $r$, as described above in Section \ref{Sec:Spatial}.
\item {\bf Enforce HSE}$[\rho_0]\rightarrow[p_0]$ computes the base state pressure, $p_0$, from a base state density, $\rho_0$ by integrating the hydrostatic equilibrium condition in one dimension. 
This follows equation (A10) in Paper V, noting that for the irregularly spaced base state case, $\Delta r$ is not constant, where $\Delta r_{j+1/2} = r_{j+1} - r_j$ for cell face with index $j+1/2$.  
The base state pressure remains equal to a constant value at the location of a prescribed cutoff density outward for the entire simulation.
\item {\bf React State}$[(\rho X_k)^{\rm in},(\rho h)^{\rm in},p_0]\rightarrow[(\rho X_k)^{\rm out},(\rho h)^{\rm out},(\rho\dot\omega),(\rho\Hnuc)]$ 
uses VODE \citep{vode} to integrate the species and enthalpy due to reactions over $\Delta t/2$ by solving
\begin{equation}
\frac{dX_k}{dt} = \dot\omega_k(\rho,X_k,T); \qquad
\frac{dT}{dt} = \frac{1}{c_p}\left(-\sum_k\xi_k\dot\omega_k + H_{\rm nuc})\right)
\end{equation}
The inputs are the species, enthalpy, and base state pressure, and the outputs are the species, enthalpy, reaction rates, and nuclear energy generation rate.
See Paper III for details.
\end{itemize}
Each time step is constrained by the standard advective CFL condition,
\begin{equation}
\Delta t = \sigma^{\rm CFL} \min_i(\Delta x / U_i),
\end{equation}
where for our simulations we typically use $\sigma^{\rm CFL}\sim 0.7$ and the minimum is taken over all spatial directions over all cells.
There are additional constraints on the time step that are typically much less restrictive than the advective CFL including
the acceleration due to the buoyancy force (sometimes in effect when the velocity is approximately zero at the start of some simulations) 
and the local magnitude of the divergence constraint (to prevent too much mass evacuation from a cell in a time step); see Section 3.4 in Paper III for details.

@@

%--------------------------------------------------------------------------
% STEP 3
%--------------------------------------------------------------------------
\item[Step 3] {\em Construct a face-centered, time-centered advective velocity, $\uadvone$.}

The construction of face-centered time-centered states used to discretize the
advection terms for velocity, species, and enthalpy, are performed using
a standard multidimensional corner transport upwind approach
\citep{colella1990multidimensional,saltzman1994unsplit} with the piecewise-parabolic method (PPM)
one-dimensional tracing \citep{colella1984piecewise}.  The full details of this
Godunov advection approach for all steps in this algorithm are described 
in Appendix A of \cite{XRB_III}.
Here we use equation (\ref{eq:momentum}) to compute face-centered, time-centered velocities, $\uadvonedag$.
The $\dagger$ superscript refers to the fact that the predicted velocity field does not satisfy the divergence constraint,
\begin{equation}
\nabla \cdot \left(\beta_0^n \uadvone\right) = \beta_0^n \left[S^{\nph,\star} - \frac{1}{\gammaonebar^np_0^n}\left(\frac{\partial p_0}{\partial t}\right)^{n-\myhalf} \right].\label{eq:div1}
\end{equation}
 We project $\uadvonedag$ onto the space of velocities that satisfy the constraint to obtain $\uadvone$.
Each projection step in the algorithm involves the solution of a variable-coefficient Poisson solve using multigrid.
Note that we still employ velocity-splitting as described by equation (\ref{eq:velsplit}) for this step 
in order to enforce the appropriate behavior of the system near the edge of the star as determined by the cutoff density.
The details of this ``MAC'' projection are provided in Appendix \ref{Sec:Projection}.

@@

\item Compute the face-centered, time-centered species, $(\rho X_k)^{\nph,\pred}$,
  for the conservative update of $(\rho X_k)^{(1)}$ using a Godunov approach \citep{XRB_III}.
  As described in Paper V, for robust numerical slope limiting we predict 
  $\rho'^n=\rho^n-\rho_0^n$ and $X_k^n$ to faces
  and here we spatially interpolate $\rho_0^n$ to faces to assemble the fluxes.

@@

\item Compute the face-centered, time-centered enthalpy, $(\rho h)^{\nph,\pred},$
  for the conservative update of $(\rho h)^{(1)}$ using using a Godunov approach \citep{XRB_III}.
  As described in Paper V, for robust numerical slope limiting 
  we predict $(\rho h)'^n=(\rho h)^n-(\rho h)_0^n$ to faces,
  where $(\rho h)_0^n$ is obtained by calling {\bf Average}$[(\rho h)^n]\rightarrow[(\rho h)_0]$,
  and here we spatially interpolate $(\rho h)_0^n$ to faces to assemble the fluxes.

@@

%--------------------------------------------------------------------------
% STEP 7
%--------------------------------------------------------------------------
\item[Step 7] {\em Construct a face-centered, time-centered advective velocity, $\uadvtwo$.}

@@

\item Compute the face-centered, time-centered species, $(\rho X_k)^{\nph,\pred}$,
  for the conservative update of $(\rho X_k)^{(1)}$ using a Godunov approach \citep{XRB_III}.
  Again, we predict $\rho'^n=\rho^n-\rho_0^n$ and $X_k^n$ to faces
  and here we spatially interpolate $(\rho_0^n+\rho_0^{n+1,*})/2$ to faces to assemble the fluxes.

@@

\item Compute the face-centered, time-centered enthalpy, $(\rho h)^{\nph,\pred},$
  for the conservative update of $(\rho h)^{(1)}$ using a Godunov approach \citep{XRB_III}.
  Again, we predict $(\rho h)'^n=(\rho h)^n-(\rho h)_0^n$ to faces
  and here we spatially interpolate $[(\rho h)_0^n)+(\rho h)_0^{n+1,*}]/2$ to faces to assemble the fluxes.

@@

First, we compute the face-centered, time-centered velocities, $\Ub^{\nph,\pred}$
using a Godunov approach \citep{XRB_III}. Then, we define
\begin{equation}
\rho^\nph = \frac{\rho^n + \rho^{n+1}}{2}, \qquad \rho_0^\nph = \frac{\rho_0^n + \rho_0^{n+1}}{2}.
\end{equation}
We update the velocity field $\Ub^n$ to $\Ub^{n+1,\dagger}$ by discretizing
equation (\ref{eq:momentum}) as
\begin{equation}
\Ub^{n+1,\dagger}
= \Ub^n - \dt \left[\uadvtwo \cdot \nabla \Ub^{\nph,\pred} \right]
 - \dt \left[ \frac{\beta_0^\nph}{\rho^\nph} \nabla \left( \frac{\pi^\nmh}{\beta_0^\nmh} \right) + \frac{\left(\rho^\nph-\rho_0^\nph\right)}{\rho^\nph} g^{\nph} \eb_r \right],
\end{equation}
Again, the $\dagger$ superscript refers
to the fact that the updated velocity does not satisfy the divergence constraint,
\begin{equation}
\nabla \cdot \left(\beta_0^{n+1} \Ub^{n+1} \right) = \beta_0^{n+1} \left[ S^{n+1} - \frac{1}{\gammaonebar^{n+1}p_0^{n+1}}\left(\frac{\partial p_0}{\partial t}\right)^{\nph}\right].\label{eq:div3}
\end{equation}
We use an approximate projection to project $\Ub^{n+1,\dagger}$ onto the space of velocities that satisfy the constraint to obtain $\Ub^{n+1}$ using a ``nodal'' projection.
This projection necessarily differs from the MAC projection used in
{\bf Step 3} and {\bf Step 7} because the velocities in those steps are defined
on faces and $\Ub^{n+1}$ is defined at cell centers, requiring different divergence
and gradient operators.
Furthermore, as part of the nodal projection, we also define a nodal new-time perturbational pressure, $\pi^\nph$.
Refer to Appendix \ref{Sec:Projection} for more details.
